---
title: "Fitabase Cleaning"
author: "Christopher Oill"
date: "2024-03-18"
output: html_document
---

# Introduction
To get my Google Data Analytics Certificate, I had to complete a case study as a final assignment. For this case study, the company "BellaBeat" requested recommendations for how to improve their products, and suggested using the "FitBit Fitness Tracker Data" data set as proxy data to aid in developing insights. The requests were generally vague, and since the requests were an imaginary scenario, I couldn't discuss more specific goals with the stakeholders. Therefore, I decided to approach the project in a more exploratory manner. As I explored, I noticed that the data provided needed some heavy cleaning, so I shifted my project focus from analysis to cleaning.

# Datasets used
The "FitBit Fitness Tracker Data" data set can be accessed [here](https://www.kaggle.com/datasets/arashnic/fitbit).

Descriptions of the columns in the data set are available [here](https://www.fitabase.com/media/1546/fitabasedatadictionary.pdf).

# Preparation
First, I loaded the libraries I would be using for this cleaning. 

```{r load libraries, include=FALSE}
library(dplyr)
library(lubridate)
library(tidyr)
library(hms)
```

Then, I uploaded the FitBit data frames to RStudio and organized them into a list. 

```{r upload data}
path_to_data <- "Fitabase Data 4.12.16-5.12.16/"
list_of_files <- list.files(path_to_data, full.names = TRUE)
fitbit_data <- lapply(list_of_files, read.csv)
```

This changed the names of each files to numbers, so I changed the names back to the names or the original files, minus ".csv".

```{r change names} 
fitbit_data <- setNames(fitbit_data, tools::file_path_sans_ext(basename(list_of_files)))
```
```{r show change names, echo=FALSE}
names(fitbit_data)
```

# High Level Cleaning
I believed having "_merged" in every title was uninformative and unnecessary, so I removed it from all titles.

```{r clean titles}
names(fitbit_data) <- sub("_merged$", "", names(fitbit_data))
```
```{r show clean titles, echo=FALSE}
names(fitbit_data)
```

After some research online, I learned that most analysts agree on titles being all lowercase, with underscores to separate words, so I adjusted all data frame titles accordingly. Then I deleted the data frames with the old titles.

```{r name cleaning}
names(fitbit_data) <- tolower(names(fitbit_data))
fitbit_data$daily_activity <- fitbit_data$dailyactivity
fitbit_data$minute_sleep <- fitbit_data$minutesleep
fitbit_data$sleep_day <- fitbit_data$sleepday
fitbit_data$weight_log_info <- fitbit_data$weightloginfo

fitbit_data$dailycalories <- NULL
fitbit_data$dailyintensities <- NULL
fitbit_data$dailysteps <- NULL
fitbit_data$dailyactivity <- NULL
fitbit_data$minutesleep <- NULL
fitbit_data$sleepday <- NULL
fitbit_data$weightloginfo <- NULL
```
```{r show clean names, echo=FALSE}
names(fitbit_data)
```

Notice that I left some of the titles alone. Many of the data frames could be cleanly consolidated into one data frame, due to matching number of rows, primary keys, and secondary keys. The next few code blocks take care of this consolidation and adds the new title formatting.

```{r consolidation}
hourly_activity <- merge(fitbit_data$hourlycalories, fitbit_data$hourlyintensities)
hourly_activity <- merge(hourly_activity, fitbit_data$hourlysteps)
fitbit_data[["hourly_activity"]] <- hourly_activity
fitbit_data$hourlycalories <- NULL
fitbit_data$hourlyintensities <- NULL
fitbit_data$hourlysteps <- NULL
hourly_activity <- NULL

minute_activity_narrow <- merge(fitbit_data$minutecaloriesnarrow, fitbit_data$minuteintensitiesnarrow)
minute_activity_narrow <- merge(minute_activity_narrow, fitbit_data$minutemetsnarrow)
minute_activity_narrow <- merge(minute_activity_narrow, fitbit_data$minutestepsnarrow)
fitbit_data[["minute_activity_narrow"]] <- minute_activity_narrow
fitbit_data$minutecaloriesnarrow <- NULL
fitbit_data$minuteintensitiesnarrow <- NULL
fitbit_data$minutemetsnarrow <- NULL
fitbit_data$minutestepsnarrow <- NULL
minute_activity_narrow <- NULL

minute_activity_wide <- merge(fitbit_data$minutecalorieswide, fitbit_data$minuteintensitieswide)
minute_activity_wide <- merge(minute_activity_wide, fitbit_data$minutestepswide)
fitbit_data[["minute_activity_wide"]] <- minute_activity_wide
fitbit_data$minutecalorieswide <- NULL
fitbit_data$minuteintensitieswide <- NULL
fitbit_data$minutestepswide <- NULL
minute_activity_wide <- NULL
```
```{r show name cleaning, echo=FALSE}
names(fitbit_data)
```

Calling back to a previous statement, most analysts agree on lower case with underscores between words for titles, so I decided now would be a good time in the code to adjust the column titles. 

```{r rename column titles}
names(fitbit_data$daily_activity) <- c("id", "date", "total_steps", "total_distance", "tracker_distance", "logged_activities_distance", "very_active_distance", "moderately_active_distance", "light_active_distance", "sedentary_active_distance", "very_active_minutes", "fairly_active_minutes", "lightly_active_minutes", "sedentary_minutes", "calories")

names(fitbit_data$heartrate_seconds) <- c("id", "date", "value")

names(fitbit_data$minute_sleep) <- c("id", "date", "value", "log_id")

names(fitbit_data$sleep_day) <- c("id", "date", "total_sleep_records", "total_minutes_asleep", "total_time_in_bed")

names(fitbit_data$weight_log_info) <- c("id", "date", "weight_kg", "weight_pounds", "fat", "bmi", "is_manual_report", "log_id")

names(fitbit_data$hourly_activity) <- c("id", "date", "calories", "total_intensity", "average_intensity", "step_total")

names(fitbit_data$minute_activity_narrow) <- c("id", "date", "calories", "intensity", "mets", "steps")

names(fitbit_data$minute_activity_wide) <- c("id", "date", "calories_00", "calories_01", "calories_02", "calories_03", "calories_04", "calories_05", "calories_06", "calories_07", "calories_08", "calories_09", "calories_10", "calories_11", "calories_12", "calories_13", "calories_14", "calories_15", "calories_16", "calories_17", "calories_18", "calories_19", "calories_20", "calories_21", "calories_22", "calories_23", "calories_24", "calories_25", "calories_26", "calories_27", "calories_28", "calories_29", "calories_30", "calories_31", "calories_32", "calories_33", "calories_34", "calories_35", "calories_36", "calories_37", "calories_38", "calories_39", "calories_40", "calories_41", "calories_42", "calories_43", "calories_44", "calories_45", "calories_46", "calories_47", "calories_48", "calories_49", "calories_50", "calories_51", "calories_52", "calories_53", "calories_54", "calories_55", "calories_56", "calories_57", "calories_58", "calories_59", "intensity_00", "intensity_01", "intensity_02", "intensity_03", "intensity_04", "intensity_05", "intensity_06", "intensity_07", "intensity_08", "intensity_09", "intensity_10", "intensity_11", "intensity_12", "intensity_13", "intensity_14", "intensity_15", "intensity_16", "intensity_17", "intensity_18", "intensity_19", "intensity_20", "intensity_21", "intensity_22", "intensity_23", "intensity_24", "intensity_25", "intensity_26", "intensity_27", "intensity_28", "intensity_29", "intensity_30", "intensity_31", "intensity_32", "intensity_33", "intensity_34", "intensity_35", "intensity_36", "intensity_37", "intensity_38", "intensity_39", "intensity_40", "intensity_41", "intensity_42", "intensity_43", "intensity_44", "intensity_45", "intensity_46", "intensity_47", "intensity_48", "intensity_49", "intensity_50", "intensity_51", "intensity_52", "intensity_53", "intensity_54", "intensity_55", "intensity_56", "intensity_57", "intensity_58", "intensity_59", "steps_00", "steps_01", "steps_02", "steps_03", "steps_04", "steps_05", "steps_06", "steps_07", "steps_08", "steps_09", "steps_10", "steps_11", "steps_12", "steps_13", "steps_14", "steps_15", "steps_16", "steps_17", "steps_18", "steps_19", "steps_20", "steps_21", "steps_22", "steps_23", "steps_24", "steps_25", "steps_26", "steps_27", "steps_28", "steps_29", "steps_30", "steps_31", "steps_32", "steps_33", "steps_34", "steps_35", "steps_36", "steps_37", "steps_38", "steps_39", "steps_40", "steps_41", "steps_42", "steps_43", "steps_44", "steps_45", "steps_46", "steps_47", "steps_48", "steps_49", "steps_50", "steps_51", "steps_52", "steps_53", "steps_54", "steps_55", "steps_56", "steps_57", "steps_58", "steps_59")
```
```{r view columns, echo=FALSE}
lapply(fitbit_data, names)
```

I noticed some clutter in my RStudio global environment, so I decided to remove it for a cleaner workspace.

```{r clean workspace}
rm(minute_activity_narrow)
rm(minute_activity_wide)
```

# Deeper Cleaning

#### Sanity Checks
Most of this data was computer generated so I expected few errors typically caused by human error, but I decided to check, anyway. I check for the following:
- duplicate rows
- extra spaces
- typos
- incorrect data types
- general inconsistencies

Unfortunately, I decided to write this markdown file after completing the project, and I didn't have the foresight at the time to keep records of my code for these sanity checks, since, as expected, I found no errors.

#### Code
Once all tables were organized, it was easier to notice some specific details I wanted to clean. The first things I noticed was the unnecessarily complex ids for each participant in the FitBit survey. I assigned sequential numbers starting at 1 to each participant, making it easier to recognize what information belonged to what participant. I should clarify that before I assigned numbers, I performed a quick unique() function on all data frames to make sure I accounted for all unique id numbers.

Since I had to do this operation on multiple data frames, and the code involved a lot of information, I made a function to save some time. 

```{r id simplification}
id_replace <- function(column_name) {sub(1503960366, 1, sub(1624580081, 2, sub(1644430081, 3, sub(1844505072, 4, sub(1927972279, 5, sub(2022484408, 6, sub(2026352035, 7, sub(2320127002, 8, sub(2347167796, 9, sub(2873212765, 10, sub(3372868164, 11, sub(3977333714, 12, sub(4020332650, 13, sub(4057192912, 14, sub(4319703577, 15, sub(4388161847, 16, sub(4445114986, 17, sub(4558609924, 18, sub(4702921684, 19, sub(5553957443, 20, sub(5577150313, 21, sub(6117666160, 22, sub(6290855005, 23, sub(6775888955, 24, sub(6962181067, 25, sub(7007744171, 26, sub(7086361926, 27, sub(8053475328, 28, sub(8253242879, 29, sub(8378563200, 30, sub(8583815059, 31, sub(8792009665, 32, sub(8877689391, 33, column_name)))))))))))))))))))))))))))))))))
}
```

I then applied this function to each data frame, to replace the old id numbers with the new ones.

```{r id replacements}
fitbit_data$daily_activity$id <- id_replace(fitbit_data$daily_activity$id)
fitbit_data$heartrate_seconds$id <- id_replace(fitbit_data$heartrate_seconds$id)
fitbit_data$minute_sleep$id <- id_replace(fitbit_data$minute_sleep$id)
fitbit_data$sleep_day$id <- id_replace(fitbit_data$sleep_day$id)
fitbit_data$weight_log_info$id <- id_replace(fitbit_data$weight_log_info$id)
fitbit_data$hourly_activity$id <- id_replace(fitbit_data$hourly_activity$id)
fitbit_data$minute_activity_narrow$id <- id_replace(fitbit_data$minute_activity_narrow$id)
fitbit_data$minute_activity_wide$id <- id_replace(fitbit_data$minute_activity_wide$id)
```

Here's an example, showing the unique values in the id column of the daily_activity data frame:

```{r example, echo=FALSE}
unique(fitbit_data$daily_activity$id)
```

The next column I decided to change was the datetime columns. When analyzing data, I personally find it more difficult to work with dates and times combined, so I split the values into separate columns. In the original columns, time was in standard time format, with AM and PM. I considered transforming this into military time, but I wasn't sure how important this detail was, so I decided to leave it alone.

```{r split datetime}
fitbit_data$minute_sleep <- mutate(separate(fitbit_data$minute_sleep, date, c("date", "time", "am_pm"), sep = " "))

fitbit_data$heartrate_seconds <- mutate(separate(fitbit_data$heartrate_seconds, date, c("date", "time", "am_pm"), sep = " "))

fitbit_data$sleep_day <- mutate(separate(fitbit_data$sleep_day, date, c("date", "time", "am_pm"), sep = " "))

fitbit_data$weight_log_info <- mutate(separate(fitbit_data$weight_log_info, date, c("date", "time", "am_pm"), sep = " "))

fitbit_data$hourly_activity <- mutate(separate(fitbit_data$hourly_activity, date, c("date", "time", "am_pm"), sep = " "))

fitbit_data$minute_activity_narrow <- mutate(separate(fitbit_data$minute_activity_narrow, date, c("date", "time", "am_pm"), sep = " "))

fitbit_data$minute_activity_wide <- mutate(separate(fitbit_data$minute_activity_wide, date, c("date", "time", "am_pm"), sep = " "))
```

Here's what the column headers look like now:

```{r view split datetime, echo=FALSE}
lapply(fitbit_data, names)
```

For some analyses, I thought that having the option to sort or filter by weekdays might be useful. For example, it might be nice to say "people ran the most on Wednesday evenings." For this reason, I added a separate column to each data frame, where I converted the date into weekdays.

```{r add weekdays column}
fitbit_data$daily_activity <- mutate(fitbit_data$daily_activity, dotw = weekdays(as.Date(fitbit_data$daily_activity$date, '%m/%d/%Y')))

fitbit_data$daily_activity <- fitbit_data$daily_activity %>% 
  relocate(dotw, .after = date)

fitbit_data$heartrate_seconds <- mutate(fitbit_data$heartrate_seconds, dotw = weekdays(as.Date(fitbit_data$heartrate_seconds$date, '%m/%d/%Y')))

fitbit_data$heartrate_seconds <- fitbit_data$heartrate_seconds %>% 
  relocate(dotw, .after = date)

fitbit_data$minute_sleep <- mutate(fitbit_data$minute_sleep, dotw = weekdays(as.Date(fitbit_data$minute_sleep$date, '%m/%d/%Y')))

fitbit_data$minute_sleep <- fitbit_data$minute_sleep %>% 
  relocate(dotw, .after = date)

fitbit_data$sleep_day <- mutate(fitbit_data$sleep_day, dotw = weekdays(as.Date(fitbit_data$sleep_day$date, '%m/%d/%Y')))

fitbit_data$sleep_day <- fitbit_data$sleep_day %>% 
  relocate(dotw, .after = date)

fitbit_data$weight_log_info <- mutate(fitbit_data$weight_log_info, dotw = weekdays(as.Date(fitbit_data$weight_log_info$date, '%m/%d/%Y')))

fitbit_data$weight_log_info <- fitbit_data$weight_log_info %>% 
  relocate(dotw, .after = date)

fitbit_data$hourly_activity <- mutate(fitbit_data$hourly_activity, dotw = weekdays(as.Date(fitbit_data$hourly_activity$date, '%m/%d/%Y')))

fitbit_data$hourly_activity <- fitbit_data$hourly_activity %>% 
  relocate(dotw, .after = date)

fitbit_data$minute_activity_narrow <- mutate(fitbit_data$minute_activity_narrow, dotw = weekdays(as.Date(fitbit_data$minute_activity_narrow$date, '%m/%d/%Y')))

fitbit_data$minute_activity_narrow <- fitbit_data$minute_activity_narrow %>% 
  relocate(dotw, .after = date)

fitbit_data$minute_activity_wide <- mutate(fitbit_data$minute_activity_wide, dotw = weekdays(as.Date(fitbit_data$minute_activity_wide$date, '%m/%d/%Y')))

fitbit_data$minute_activity_wide <- fitbit_data$minute_activity_wide %>% 
  relocate(dotw, .after = date)
```

Here's an example, using the weight_log_info data frame:

```{r weekday example, echo=FALSE}
glimpse(fitbit_data$weight_log_info)
```

After I put new columns into the data frames, I noticed that the data types were all "character" types. I wanted the type to be more representative of the information in each column, so I changed each data type appropriately, turning id numbers into integers, date into a yyyy-mm-dd date type, and time into hh:mm:ss format. I also changed a True or False column that wasn't listed as Boolean type.

```{r change data types}
fitbit_data$heartrate_seconds$id <- as.integer(fitbit_data$heartrate_seconds$id)
fitbit_data$heartrate_seconds$date <- as.Date(fitbit_data$heartrate_seconds$date, '%m/%d/%Y')
fitbit_data$heartrate_seconds$time <- as_hms(fitbit_data$heartrate_seconds$time)

fitbit_data$daily_activity$id <- as.integer(fitbit_data$daily_activity$id)
fitbit_data$daily_activity$date <- as.Date(fitbit_data$daily_activity$date, '%m/%d/%Y')

fitbit_data$minute_sleep$id <- as.integer(fitbit_data$minute_sleep$id)
fitbit_data$minute_sleep$date <- as.Date(fitbit_data$minute_sleep$date, '%m/%d/%Y')
fitbit_data$minute_sleep$time <- as_hms(fitbit_data$minute_sleep$time)

fitbit_data$sleep_day$id <- as.integer(fitbit_data$sleep_day$id)
fitbit_data$sleep_day$date <- as.Date(fitbit_data$sleep_day$date, '%m/%d/%Y')
fitbit_data$sleep_day$time <- as_hms(fitbit_data$sleep_day$time)

fitbit_data$weight_log_info$id <- as.integer(fitbit_data$weight_log_info$id)
fitbit_data$weight_log_info$date <- as.Date(fitbit_data$weight_log_info$date, '%m/%d/%Y')
fitbit_data$weight_log_info$time <- as_hms(fitbit_data$weight_log_info$time)
fitbit_data$weight_log_info$is_manual_report <- as.logical(fitbit_data$weight_log_info$is_manual_report)

fitbit_data$hourly_activity$id <- as.integer(fitbit_data$hourly_activity$id)
fitbit_data$hourly_activity$date <- as.Date(fitbit_data$hourly_activity$date, '%m/%d/%Y')
fitbit_data$hourly_activity$time <- as_hms(fitbit_data$hourly_activity$time)

fitbit_data$minute_activity_narrow$id <- as.integer(fitbit_data$minute_activity_narrow$id)
fitbit_data$minute_activity_narrow$date <- as.Date(fitbit_data$minute_activity_narrow$date, '%m/%d/%Y')
fitbit_data$minute_activity_narrow$time <- as_hms(fitbit_data$minute_activity_narrow$time)

fitbit_data$minute_activity_wide$id <- as.integer(fitbit_data$minute_activity_wide$id)
fitbit_data$minute_activity_wide$date <- as.Date(fitbit_data$minute_activity_wide$date, '%m/%d/%Y')
fitbit_data$minute_activity_wide$time <- as_hms(fitbit_data$minute_activity_wide$time)
```

Here's an example, using the weight_log_info data frame again.

```{r data type example, echo=FALSE}
glimpse(fitbit_data$weight_log_info)
```

For analysis purposes, the log_id column was pointless, as it was just a number assigned to each separate date. I imagine it has a purpose within FitBit's inner workings, but with analysis, it's more useful to just use the date itself. I removed this column in all data frames which contained it.

```{r remove log_id column}
fitbit_data$minute_sleep$log_id <- NULL
fitbit_data$weight_log_info$log_id <- NULL
```

I created a folder named "Fitabase_Data_4.12.16-5.12.16_Cleaned" for the data frames to be exported to.

```{r folder creation}
dir.create("Fitabase_Data_4.12.16-5.12.16_Cleaned")
```

To prepare the cleaned data for upload to other websites, I exported each data frame separately into the folder as csv files.

```{r export new frames}
for(framename in names(fitbit_data)) {
  write.csv(fitbit_data[[framename]], paste0("Fitabase_Data_4.12.16-5.12.16_Cleaned/", framename, ".csv"), row.names = FALSE)
}
```

# Second Data Set

As I was uploading my cleaned data set to Kaggle, I needed to check the original data set owner's page for some metadata information. I noticed that a second data set had been added since I started working with the original one. This new data set covered the same participants, but from the month before. 

On further investigation, I noticed a few differences between the two sets, which would affect the necessary code to perform a cleaning. First, many participants didn't start until a few weeks into the first month. Second, a couple of participants were in the first month, but not the second month. Third, a few data tables were in the second month that weren't in the first month.

I imagine the owner of the data sets originally left out the first month due to its inconsistencies and lack of data. However, I decided to perform the same cleaning processes on this data set, despite it's deficiencies, as I believe it can still contribute to analyses.

Here is the code for the whole cleaning. Notice that any "sleep_day" and "minute_activity_wide" data frames have been removed, and participant 34 and 35 have been added to any relevant code.

```{r second set cleaning}
# import multiple data frames into one list with original names
path_to_data2 <- "Fitabase Data 3.12.16-4.11.16/"
list_of_files2 <- list.files(path_to_data2, full.names = TRUE)
fitbit_data2 <- lapply(list_of_files2, read.csv)
fitbit_data2 <- setNames(fitbit_data2, tools::file_path_sans_ext(basename(list_of_files2)))

# substitute names in fitbit_data to remove "_merged"
names(fitbit_data2) <- sub("_merged$", "", names(fitbit_data2))

# make all titles lowercase and add underscores
names(fitbit_data2) <- tolower(names(fitbit_data2))
fitbit_data2$daily_activity <- fitbit_data2$dailyactivity
fitbit_data2$minute_sleep <- fitbit_data2$minutesleep
fitbit_data2$weight_log_info <- fitbit_data2$weightloginfo

# removes data frames with repeated data and data frames with old titles
fitbit_data2$dailycalories <- NULL
fitbit_data2$dailyintensities <- NULL
fitbit_data2$dailysteps <- NULL
fitbit_data2$dailyactivity <- NULL
fitbit_data2$minutesleep <- NULL
fitbit_data2$weightloginfo <- NULL

# combines all hourly data frames into one data frame
hourly_activity <- merge(fitbit_data2$hourlycalories, fitbit_data2$hourlyintensities)
hourly_activity <- merge(hourly_activity, fitbit_data2$hourlysteps)
fitbit_data2[["hourly_activity"]] <- hourly_activity
fitbit_data2$hourlycalories <- NULL
fitbit_data2$hourlyintensities <- NULL
fitbit_data2$hourlysteps <- NULL
hourly_activity <- NULL

# combines all narrow minute data frames into one data frame
minute_activity_narrow <- merge(fitbit_data2$minutecaloriesnarrow, fitbit_data2$minuteintensitiesnarrow)
minute_activity_narrow <- merge(minute_activity_narrow, fitbit_data2$minutemetsnarrow)
minute_activity_narrow <- merge(minute_activity_narrow, fitbit_data2$minutestepsnarrow)
fitbit_data2[["minute_activity_narrow"]] <- minute_activity_narrow
fitbit_data2$minutecaloriesnarrow <- NULL
fitbit_data2$minuteintensitiesnarrow <- NULL
fitbit_data2$minutemetsnarrow <- NULL
fitbit_data2$minutestepsnarrow <- NULL
minute_activity_narrow <- NULL

# rename columns with lowercase and underscores
names(fitbit_data2$daily_activity) <- c("id", "date", "total_steps", "total_distance", "tracker_distance", "logged_activities_distance", "very_active_distance", "moderately_active_distance", "light_active_distance", "sedentary_active_distance", "very_active_minutes", "fairly_active_minutes", "lightly_active_minutes", "sedentary_minutes", "calories")

names(fitbit_data2$heartrate_seconds) <- c("id", "date", "value")

names(fitbit_data2$minute_sleep) <- c("id", "date", "value", "log_id")

names(fitbit_data2$weight_log_info) <- c("id", "date", "weight_kg", "weight_pounds", "fat", "bmi", "is_manual_report", "log_id")

names(fitbit_data2$hourly_activity) <- c("id", "date", "calories", "total_intensity", "average_intensity", "step_total")

names(fitbit_data2$minute_activity_narrow) <- c("id", "date", "calories", "intensity", "mets", "steps")

# remove values from global environment
rm(hourly_activity)
rm(minute_activity_narrow)

# create a function to replace ids to ease readability
id_replace <- function(column_name) {sub(1503960366, 1, sub(1624580081, 2, sub(1644430081, 3, sub(1844505072, 4,  sub(1927972279, 5, sub(2022484408, 6, sub(2026352035, 7, sub(2320127002, 8, sub(2347167796, 9, sub(2873212765, 10, sub(3372868164, 11, sub(3977333714, 12, sub(4020332650, 13, sub(4057192912, 14, sub(4319703577, 15, sub(4388161847, 16, sub(4445114986, 17, sub(4558609924, 18, sub(4702921684, 19, sub(5553957443, 20, sub(5577150313, 21, sub(6117666160, 22, sub(6290855005, 23, sub(6775888955, 24, sub(6962181067, 25, sub(7007744171, 26, sub(7086361926, 27, sub(8053475328, 28, sub(8253242879, 29, sub(8378563200, 30, sub(8583815059, 31, sub(8792009665, 32, sub(8877689391, 33, sub(2891001357, 34, sub(6391747486, 35, column_name)))))))))))))))))))))))))))))))))))
}

# execute id_replace function on each data frame
fitbit_data2$daily_activity$id <- id_replace(fitbit_data2$daily_activity$id)
fitbit_data2$heartrate_seconds$id <- id_replace(fitbit_data2$heartrate_seconds$id)
fitbit_data2$minute_sleep$id <- id_replace(fitbit_data2$minute_sleep$id)
fitbit_data2$weight_log_info$id <- id_replace(fitbit_data2$weight_log_info$id)
fitbit_data2$hourly_activity$id <- id_replace(fitbit_data2$hourly_activity$id)
fitbit_data2$minute_activity_narrow$id <- id_replace(fitbit_data2$minute_activity_narrow$id)

# Organization ----

# separates dates and times into two columns
fitbit_data2$minute_sleep <- mutate(separate(fitbit_data2$minute_sleep, date, c("date", "time", "am_pm"), sep = " "))

fitbit_data2$heartrate_seconds <- mutate(separate(fitbit_data2$heartrate_seconds, date, c("date", "time", "am_pm"), sep = " "))

fitbit_data2$weight_log_info <- mutate(separate(fitbit_data2$weight_log_info, date, c("date", "time", "am_pm"), sep = " "))

fitbit_data2$hourly_activity <- mutate(separate(fitbit_data2$hourly_activity, date, c("date", "time", "am_pm"), sep = " "))

fitbit_data2$minute_activity_narrow <- mutate(separate(fitbit_data2$minute_activity_narrow, date, c("date", "time", "am_pm"), sep = " "))

# convert dates to weekdays, then add as a separate column to each data frame
fitbit_data2$daily_activity <- mutate(fitbit_data2$daily_activity, dotw = weekdays(as.Date(fitbit_data2$daily_activity$date, '%m/%d/%Y')))

fitbit_data2$daily_activity <- fitbit_data2$daily_activity %>% 
  relocate(dotw, .after = date)

fitbit_data2$heartrate_seconds <- mutate(fitbit_data2$heartrate_seconds, dotw = weekdays(as.Date(fitbit_data2$heartrate_seconds$date, '%m/%d/%Y')))

fitbit_data2$heartrate_seconds <- fitbit_data2$heartrate_seconds %>% 
  relocate(dotw, .after = date)

fitbit_data2$minute_sleep <- mutate(fitbit_data2$minute_sleep, dotw = weekdays(as.Date(fitbit_data2$minute_sleep$date, '%m/%d/%Y')))

fitbit_data2$minute_sleep <- fitbit_data2$minute_sleep %>% 
  relocate(dotw, .after = date)

fitbit_data2$weight_log_info <- mutate(fitbit_data2$weight_log_info, dotw = weekdays(as.Date(fitbit_data2$weight_log_info$date, '%m/%d/%Y')))

fitbit_data2$weight_log_info <- fitbit_data2$weight_log_info %>% 
  relocate(dotw, .after = date)

fitbit_data2$hourly_activity <- mutate(fitbit_data2$hourly_activity, dotw = weekdays(as.Date(fitbit_data2$hourly_activity$date, '%m/%d/%Y')))

fitbit_data2$hourly_activity <- fitbit_data2$hourly_activity %>% 
  relocate(dotw, .after = date)

fitbit_data2$minute_activity_narrow <- mutate(fitbit_data2$minute_activity_narrow, dotw = weekdays(as.Date(fitbit_data2$minute_activity_narrow$date, '%m/%d/%Y')))

fitbit_data2$minute_activity_narrow <- fitbit_data2$minute_activity_narrow %>% 
  relocate(dotw, .after = date)

# fix data types
fitbit_data2$heartrate_seconds$id <- as.integer(fitbit_data2$heartrate_seconds$id)
fitbit_data2$heartrate_seconds$date <- as.Date(fitbit_data2$heartrate_seconds$date, '%m/%d/%Y')
fitbit_data2$heartrate_seconds$time <- as_hms(fitbit_data2$heartrate_seconds$time)

fitbit_data2$daily_activity$id <- as.integer(fitbit_data2$daily_activity$id)
fitbit_data2$daily_activity$date <- as.Date(fitbit_data2$daily_activity$date, '%m/%d/%Y')

fitbit_data2$minute_sleep$id <- as.integer(fitbit_data2$minute_sleep$id)
fitbit_data2$minute_sleep$date <- as.Date(fitbit_data2$minute_sleep$date, '%m/%d/%Y')
fitbit_data2$minute_sleep$time <- as_hms(fitbit_data2$minute_sleep$time)

fitbit_data2$weight_log_info$id <- as.integer(fitbit_data2$weight_log_info$id)
fitbit_data2$weight_log_info$date <- as.Date(fitbit_data2$weight_log_info$date, '%m/%d/%Y')
fitbit_data2$weight_log_info$time <- as_hms(fitbit_data2$weight_log_info$time)
fitbit_data2$weight_log_info$is_manual_report <- as.logical(fitbit_data2$weight_log_info$is_manual_report)

fitbit_data2$hourly_activity$id <- as.integer(fitbit_data2$hourly_activity$id)
fitbit_data2$hourly_activity$date <- as.Date(fitbit_data2$hourly_activity$date, '%m/%d/%Y')
fitbit_data2$hourly_activity$time <- as_hms(fitbit_data2$hourly_activity$time)

fitbit_data2$minute_activity_narrow$id <- as.integer(fitbit_data2$minute_activity_narrow$id)
fitbit_data2$minute_activity_narrow$date <- as.Date(fitbit_data2$minute_activity_narrow$date, '%m/%d/%Y')
fitbit_data2$minute_activity_narrow$time <- as_hms(fitbit_data2$minute_activity_narrow$time)

# remove unnecessary columns
fitbit_data2$minute_sleep$log_id <- NULL
fitbit_data2$weight_log_info$log_id <- NULL

# create folder for files
dir.create("Fitabase_Data_3.12.16-4.11.16_Cleaned")

# export data frames to csv files
for(framename in names(fitbit_data2)) {
  write.csv(fitbit_data2[[framename]], paste0("Fitabase_Data_3.12.16-4.11.16_Cleaned/", framename, ".csv"), row.names = FALSE)
}
```

# Merged Data Set

After cleaning both data sets, I thought it might be nice for analysts to have the option to work with both sets combined. If an analyst wanted the data split a different way, rather than by month, this would make their job a little easier.

First, I performed a full join for all data frames that each month had in common, and created new variables for the data frames only found in the second month.

```{r join data frames, include=FALSE}
daily_activity <- full_join(fitbit_data2$daily_activity, fitbit_data$daily_activity)
heartrate_seconds <- full_join(fitbit_data2$heartrate_seconds, fitbit_data$heartrate_seconds)
hourly_activity <- full_join(fitbit_data2$hourly_activity, fitbit_data$hourly_activity)
minute_activity_narrow <- full_join(fitbit_data2$minute_activity_narrow, fitbit_data$minute_activity_narrow)
minute_sleep <- full_join(fitbit_data2$minute_sleep, fitbit_data$minute_sleep)
sleep_day <- fitbit_data$sleep_day
weight_log_info <- full_join(fitbit_data2$weight_log_info, fitbit_data$weight_log_info)
minute_activity_wide <- fitbit_data$minute_activity_wide
sleep_day <- fitbit_data$sleep_day
```

I created a new list, combining all of the newly merged data frames, and the data frames that were converted into their own variables.

```{r new list}
fitbit_data_merged <- list(daily_activity, heartrate_seconds, hourly_activity, minute_activity_narrow, minute_activity_wide, minute_sleep, sleep_day, weight_log_info)
fitbit_data_merged <- setNames(fitbit_data_merged, c("daily_activity", "heartrate_seconds", "hourly_activity", "minute_activity_narrow", "minute_activity_wide", "minute_sleep", "sleep_day", "weight_log_info"))
```

For a clean workspace, I removed the data frames from the global environment.
```{r remove data frames}
rm(daily_activity)
rm(heartrate_seconds)
rm(hourly_activity)
rm(minute_activity_narrow)
rm(minute_activity_wide)
rm(minute_sleep)
rm(sleep_day)
rm(weight_log_info)
```

I created a folder for the merged files to go into, called "Fitabase_Data_3.12.16-5.12.16_Cleaned".

```{r create merged data folder}
dir.create("Fitabase_Data_3.12.16-5.12.16_Cleaned")
```

I exported the data frames into .csv files, where I then combined them into a zip file. 

```{r export merged data}
for(framename in names(fitbit_data_merged)) {
  write.csv(fitbit_data_merged[[framename]], paste0("Fitabase_Data_3.12.16-5.12.16_Cleaned/", framename, ".csv"), row.names = FALSE)
}
```

# Kaggle

After cleaning both sets, and then combining them into a new set, I uploaded all three to my Kaggle profile. You can find these data sets [here](https://www.kaggle.com/datasets/chrisoill/fitbit-fitness-tracker-data-cleaned). The original source did not have any descriptions about each separate data frame or each column, so I used the information I found from the fitabase data dictionary linked previously to help fill in this missing information. 

# Future Directions

Now that the data sets are cleaned and shared, I will perform my own exploratory analysis. Based on some interesting observations I noticed while cleaning, I suspect I may find some useful trends by inspecting moments when participants weren't using the fitbit, tables which only included optional participation, activity on certain dates, and activity during certain times. 

I hope others find this cleaned data as helpful as I will!